Semantic Linearity in a Lazy Optimising Compiler

### 2025-07-09 icfp_submission_2025.tex
save submission 2025 as separate file

We present Linear Core, the first type system that understands semantic
linearity in the presence of non-strict evaluation, suitable for the Core
intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
is sound and guarantees linear resource usage and prove multiple optimising
transformations preserve linearity in Linear Core while failing to do so in
Core. We have implemented Linear Core as a compiler plugin to validate the
system against linearity-heavy libraries, including
\texttt{linear-base}.%^, in the heart of the compiler.
\end{abstract}

@@

Both Haskell and its intermediate language Core are lazily
        evaluated.
        % -- expressions in Haskell are only evaluated when needed.
        As we make clear in this paper, \emph{non-strict evaluation interacts non-trivially with linearity}.
        Intuitively, a syntactic occurrence of a linear resource does not
        necessarily entail consuming that resource since expressions are not
        necessarily evaluated (i.e., if the expression is not evaluated, the
        resource is not used).
        %
        Under eager evaluation, a syntactic occurrence of a resource
        along a given control-flow path always implies its use at runtime.
        %
        % Under eager/call-by-value evaluation, the distinction between syntactic uses of a
        % resource and the actual use of linear resources at runtime does not exist --
        % an occurrence of a variable in the program always entails consuming it.
        %

>>>>

We present Linear Core, the first type system that understands semantic
linearity in the presence of non-strict evaluation, suitable for the Core
intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
is sound and guarantees linear resource usage, and prove multiple optimising
transformations preserve linearity in Linear Core while failing to do so in
Core. We have implemented Linear Core as a compiler plugin to validate the
system against linearity-heavy libraries, including
\texttt{linear-base}.%^, in the heart of the compiler.
\end{abstract}

@@

Both Haskell and its intermediate language Core are lazily
        evaluated.
        % -- expressions in Haskell are only evaluated when needed.
        As we make clear in this paper, \emph{non-strict evaluation interacts non-trivially with linearity}.
        Intuitively, a syntactic occurrence of a linear resource does not
        necessarily entail consuming that resource because expressions are not
        necessarily evaluated (if the expression is not evaluated, the
        resource is not used).
        %
        In contrast, under eager evaluation, a syntactic occurrence of a resource
        along a given control-flow path always implies its use at runtime.
        %
        % Under eager/call-by-value evaluation, the distinction between syntactic uses of a
        % resource and the actual use of linear resources at runtime does not exist --
        % an occurrence of a variable in the program always entails consuming it.
        %
